Effect of resonant and random excitation on the LHC beam in the context of pulsed hollow electron lens operation.

### 2018-02-14 resex.tex
Comments from Giulio for random excitation

The ADT kicker deflection angle is calculated from the kicker geometry and the excitation voltage on the deflection plates. The excitation voltage used in the experiments is a complicated waveform, generated by a very complex hardware (digital signal processor, chain of transmission lines, low and high power amplifiers). The peak excitation voltage (and subsequently the kick) seen by the beam can be only indirectly measured using the probes on each kicker. The measured values are therefore believed to be known with an error margin of 10-15\% for the experiments in 2017 and 50\% in 2016. The estimate for 2017 is based on only a single measurement and in the future and with more time available the determination of the error would definitely have to be improved. The error in 2016 is an educated guess as no calibration was performed beforehand. The maximum kick strength reachable with the ADT at injection without risking saturation is 96~nrad. As an example Fig.~\ref{fig:fill_meas} shows the oscillation amplitude of each bunch in both H and V planes captured by the LHC real time transverse activity monitor during the 2017 experiments and an excitation with $8^{\mathrm{th}}$ turn pulsing. The measured activity is well in line with the expected excitation pattern.
\begin{figure}[h]
	\centering
	\includegraphics[width=1.0\linewidth]{bunchfilling_measured.png}	
	\caption{\label{fig:fill_meas} Bunch by bunch excitation detected by the real time transverse activity monitor during the 2017 experiment and an excitation with 8th turn pulsing.}
\end{figure}

@@

The simulation results from the distribution tracking shown in Fig.~\ref{fig:7thsim} in general predict in terms of emittance for the two experiments:
\begin{description}
	\item[2016 experiment] Excitation amplitude dependent emittance growth in the horizontal but not in the vertical plane.
	\item[2017 experiment] Excitation amplitude dependent emittance growth in the vertical plane but not in the horizontal plane for an excitation in V and H+V. For an excitation only in H no emittance growth is observed.
\end{description}
Similar changes of the emittance as predicted in simulations were also experimentally observed as depicted in Fig.~\ref{fig:7thexp}, where emittance growth was observed only in the horizontal plane for pulsing in H in 2016, and only in the vertical plane for pulsing in V and H+V in 2017 with no emittance growth at all for pulsing in H. Just as for the $10^{\mathrm{th}}$~turn pulsing, the emittance growth can be divided also for the $7^{\mathrm{th}}$~turn pulsing in a phase of fast adjustment of the beam distribution and the reach of a new equilibrium state indicated in blue (fast adjustment) and black (equilibrium state) in Fig.~\ref{fig:7thexp}. Also in this case, the distribution changes are directly visible on the beam profiles taken with the BSRT (see Appendix~\ref{app:sec:7}, Fig.~\ref{fig:7thexpprof}).
\begin{figure*}[h]
	%	\begin{minipage}[t]{0.49\linewidth}
	%		\centering
	%		2016 experiment
	%		\includegraphics[width=1.0\linewidth]{2016_bunch_intensity_h7th_no_damper_avg.png}
	%	\end{minipage}	
	%	\begin{minipage}[t]{0.49\linewidth}
	%		\centering
	%		2017 experiment
	%		\includegraphics[width=1.0\linewidth]{2017_bunch_intensity_hv7th_no_damper_avg.png}
	%	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		2016 experiment, $7^\mathrm{th}$ H	\includegraphics[width=1.0\linewidth]{2016_emith_avg_rel_h7th_no_damper.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		2017 experiment, $7^\mathrm{th}$ H+V
		\includegraphics[width=1.0\linewidth]{2017_emith_avg_rel_hv7th_no_damper.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2016_emitv_avg_rel_h7th_no_damper.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2017_emitv_avg_rel_hv7th_no_damper.png}
	\end{minipage}	
	\caption{\label{fig:7thexp} Relative horizontal (top) and vertical (bottom) emittance obtained during the 2016 experiments (left) and the 2017 experiment (right) averaged over the bunches experiencing the same excitation amplitude. For all bunches the damper was not active.}
\end{figure*}

@@

\subsection{random excitation\label{sec:simexran}}
The random excitation presents in general the pulsing pattern leading to the fastest halo depletion rates achievable with an HEL. On the other hand this pattern is also the most dangerous one for the proton beam core. The aim of this experiment was to test the effect on the core in comparison to the resonant excitation represented in the 2017 experiment by primarily the $7^{\mathrm{th}}$~turn pulsing as for $8^{\mathrm{th}}$~turn pulsing basically no effect was observed. The random excitation is in general so powerful as it excites all frequencies of the beam equally.
\begin{figure}[h]
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		no excitation
		\includegraphics[width=1.0\linewidth]{2017injnocolc15o+19_6noerru_dp0_amp.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		uniform random H+V
		\includegraphics[width=1.0\linewidth]{2017injnocolc15o+19_6noerruranadthv_1nrad_dp0_amp.png}
	\end{minipage}	
	\caption{\label{fig:ran2017fmaamp} FMA analysis in amplitude space without excitation (left) and for a uniform random excitation with 1~nrad amplitude (right) in H+V based on the 2017 injection optics with no machine errors and a tune of (62.27,60.295).}
\end{figure}
This can be seen formidably on the example of the FMA analysis in amplitude space, depicted in Fig.~\ref{fig:ran2017fmaamp}, for a random excitation as applied during the 2017 experiment. In this case, the diffusion in tune is increased equally by the random excitation independent of the particle amplitude.

\begin{figure}[h]
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2017injerra2b2u_ranadt_3_5um_intensity.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2017injerra2b2u_ranadt_3_5um_sigm.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2017injerra2b2u_ranadt_3_5um_emit1.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2017injerra2b2u_ranadt_3_5um_emit2.png}
	\end{minipage}
	\caption{\label{fig:ransim} Relative intensity (top left), bunch length (top right), and horizontal (bottom left) and vertical (bottom right) emittance obtained from distribution tracking based on the 2017 injection optics with standard errors and $(Q_x,Q_y)=(62.27,60.295)$. The solid line indicates the case with no excitation, and the dotted and dashed line the results for a uniform random excitation of 12~nrad and 24~nrad respectively.}
\end{figure}
The results from the distribution tracking are shown in Fig.~\ref{fig:ransim}. In contrast to the $7^{\mathrm{th}}$ and $10^{\mathrm{th}}$~turn pulsing, the effect of a random excitation is very symmetric:
\begin{itemize}
	\item Emittance growth is only caused in the plane of excitation.
	\item The emittance growth is comparable between an excitation in H and V in the corresponding plane as well as the scaling with the excitation amplitude. Only a small increase or decrease is observed if the excitation is applied in both planes. The difference could be due to coupling.
	\item The random excitation leads to an increased constant emittance growth rate which is reached without undergoing an initial adjustment phase.
	\item Negligible losses are observed independent of the plane of excitation.
\end{itemize}

\begin{figure}[b]
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		random excitation V
		\includegraphics[width=1.0\linewidth]{2017_bunch_intensity_vran_no_damper_avg.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		random excitation H+V
		\includegraphics[width=1.0\linewidth]{2017_bunch_intensity_hvran_no_damper_avg.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2017_emith_avg_rel_vran_no_damper.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2017_emith_avg_rel_hvran_no_damper.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2017_emitv_avg_rel_vran_no_damper.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2017_emitv_avg_rel_hvran_no_damper.png}
	\end{minipage}	
	\caption{\label{fig:ranexp} Relative intensity (top) and relative horizontal (center) and vertical (bottom) emittance for a random excitation only in V (left) and in H+V (right). Emittance growth is only caused in the plane of excitation.}
\end{figure}
Qualitatively, this behavior could be also confirmed experimentally during the 2017 experiments, where excitation amplitude dependent emittance growth was only recorded in the plane of excitation with all excitation amplitude dependent emittance growth rates in a comparable range. As example the relative horizontal and vertical emittance for pulsing in V and in H+V are depicted in Fig.~\ref{fig:ranexp}. Note that as expected from simulations and despite the large excitation amplitude dependent emittance growth, the growth rate stays constant and there is explicitly no fast adjustment phase succeeded by a new equilibrium state as was observed for the $7^{\mathrm{th}}$ and $10^{\mathrm{th}}$~turn pulsing.

\begin{figure}[h]
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		reference bunch,\\ no excitation
		\includegraphics[width=1.0\linewidth]{profile_v_ranv_slot_1698.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		excited bunch,\\ random excitation V
		\includegraphics[width=1.0\linewidth]{profile_v_ranv_slot_1532.png}
	\end{minipage}
	\caption{\label{fig:ranexpprof}Vertical beam profiles measured with the Beam Synchrotron Radiation Telescope (BSRT) during the 2017 experiments. The profiles are taken at the end of the random excitation in V. For both bunches the transverse damper is not active. The distribution changes in the reference bunch (left), here slot 1698, are much less than for the bunch experiencing the maximum excitation (right), here slot 2696.}
\end{figure}
The large changes of the emittance are of course also directly visible in the beam profiles measured with the BSRT shown in Fig.~\ref{fig:ranexpprof} for a random excitation only in V. In comparison to the $10^{\mathrm{th}}$~turn pulsing shown in Fig.~\ref{fig:10thexpprof}, the distribution also stays Gaussian in case of a random excitation while for the $10^{\mathrm{th}}$~turn pulsing it clearly adopts to a non-Gaussian shape.

\begin{figure}[b]
	\centering
	\includegraphics[width=0.8\linewidth]{2017_scale_amp_ranhv_lbllong.png}
	\caption{\label{fig:ranexploss}  Comparison of scaling of loss rates with excitation amplitude for a random excitation and $7^{\mathrm{th}}$~turn pulsing in H+V measured during the 2017 experiment. The excitation amplitude error is 15\% indicated by the horizontal error bars.}
\end{figure}
The experimental results for a maximum excitation amplitude of $A_{\mathrm{max}}=1$~nrad and $A_{\mathrm{max}}=6$~nrad also confirm the in general small losses predicted in simulations. The losses then suddenly increase once the maximum excitation amplitude is increased to $A_{\mathrm{max}}=12$~nrad, which is most likely the point when, due to the large emittance increase, the physical aperture starts eating into the more central region of the distribution resulting in higher losses despite the larger emittance growth for a random excitation. In comparison to the $7^{\mathrm{th}}$~turn pulsing, the loss rate for a random excitation is smaller for a maximum excitation amplitude of $A_{\mathrm{max}}=6$~nrad, but larger for $A_{\mathrm{max}}=12$~nrad (see Fig.~\ref{fig:ranexploss}). This indicates an increase of the diffusion mostly in the beam tails for $7^{\mathrm{th}}$~turn pulsing, while for a random excitation mostly the core is affected. Furthermore, the dependence of the loss rate on the excitation plane is additive, meaning that double the loss rate is observed for an excitation in H+V compared to only in H or only V (see Appendix~\ref{app:sec:ran}, Fig.~\ref{fig:ranexplossplane}). In view of the next section, Sec.~\ref{sec:damp}, not that the loss rate is also considerably reduced for a random excitation if the transverse damper is active, while in case of the $7^{\mathrm{th}}$~turn pulsing the loss rate does not change.

\subsection{Effect of the transverse damper\label{sec:damp}}
\begin{figure}[b]
	\centering
	random excitation H+V\\
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		damper not active
		\includegraphics[width=1.0\linewidth]{2017_emitv_avg_rel_vran_no_damper.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		damper active
		\includegraphics[width=1.0\linewidth]{2017_emitv_avg_rel_vran_with_damper.png}
	\end{minipage}	
	\centering
	$7^{\mathrm{th}}$~turn H+V\\
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		damper not active
		\includegraphics[width=1.0\linewidth]{2017_emitv_avg_rel_v7th_no_damper_no_text.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		damper active
		\includegraphics[width=1.0\linewidth]{2017_emitv_avg_rel_v7th_with_damper_no_text.png}
	\end{minipage}	
	\caption{\label{fig:damp} Relative intensity for bunches with transverse damper not active (top) and transverse damper active (bottom) for $7^{\mathrm{th}}$~turn pulsing in V (left) and for a random excitation only in V (right).}
\end{figure}
For all $k^{\mathrm{th}}$~turn pulsing patterns, the transverse damper does in general not change any of the observables studied during the experiments, explicitly losses, emittance and beam distribution suggesting that indeed it is not capable of damping any resonant excitation. On the other hand, the transverse damper considerably decreases any changes of the above parameters in case of a random excitation. As an example, the vertical emittance growth for the $7^{\mathrm{th}}$~turn pulsing in V  is compared with a random excitation in V in Fig.~\ref{fig:damp} and the loss rates in Fig.~\ref{fig:ranexploss} illustrating impressively the aforesaid behavior. 
\textcolor{red}{Any idea why? Daniel, help!}

\section{Summary\label{sec:sum}}
This paper intends to give an overview of the effects of the HEL on the proton beam core and the experiments conducted in this context at the LHC in 2016~\cite{resexmd2016} and 2017~\cite{resexmd2017}. For a perfect HEL without any imperfections no effect on the core is to be expected as the field at the center vanishes by construction. Only in the case of imperfections in the guidance and profile of the electron beam, a residual kick on the proton beam core can occur. In this case and assuming HL-LHC parameters (Tables~\ref{tab:hllhc_param}--\ref{tab:hel_param}), the contribution from the electron lens bends with approximately $0.5$~nrad to first order is negligible with respect to the kick expected from the central region of the HEL of approximatley $20$~nrad to first order caused by profile imperfections in the electron beam.

@@

\section{Additional plots for random excitation during 2017 experiments}
\label{app:sec:ran}
\begin{figure}[h]
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2017_scale_amp_ranh_lbllong.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2017_scale_amp_ranv_lbllong.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2017_scale_amp_ranhv_lbllong.png}
	\end{minipage}	
	\caption{\label{fig:ranexplossplane} Comparison of scaling of loss rates with excitation amplitude for a random excitation and $7^{\mathrm{th}}$~turn pulsing only in H (top), only in V (center) and in H+V (bottom) measured during the 2017 experiment. The excitation amplitude error is 15\% indicated by the horizontal error bars. Note that for $7^{\mathrm{th}}$~turn pulsing the transverse damper does not influence the loss rate, while for a random excitation the loss rate is considerably reduced for the bunches with transverse damper active.}
\end{figure}

>>>>

The ADT kicker deflection angle is calculated from the kicker geometry and the excitation voltage on the deflection plates. The excitation voltage used in the experiments is a complicated waveform, generated by a very complex hardware (digital signal processor, chain of transmission lines, low and high power amplifiers). The peak excitation voltage (and subsequently the kick) seen by the beam can be only indirectly measured using the probes on each kicker. The measured values are therefore believed to be known with an error margin of 10-15\% for the experiments in 2017 and 50\% in 2016. The estimate for 2017 is based on only a single measurement and in the future and with more time available the determination of the error would definitely have to be improved. The error in 2016 is estimated from the transverse damper operational parameters. Due to lack of the dedicated machine time no precision calibration was performed. The maximum kick strength reachable with the ADT at injection without risking saturation is 96~nrad. As an example Fig.~\ref{fig:fill_meas} shows the oscillation amplitude of each bunch in both H and V planes captured by the LHC real time transverse activity monitor during the 2017 experiments and an excitation with $8^{\mathrm{th}}$ turn pulsing. The measured activity is well in line with the expected excitation pattern.
\begin{figure}[h]
	\centering
	\includegraphics[width=1.0\linewidth]{bunchfilling_measured.png}	
	\caption{\label{fig:fill_meas} Bunch by bunch excitation detected by the real time transverse activity monitor during the 2017 experiment and an excitation with 8th turn pulsing.}
\end{figure}

@@

The simulation results from the distribution tracking shown in Fig.~\ref{fig:7thsim} in general predict in terms of emittance for the two experiments:
\begin{description}
	\item[2016 experiment] Excitation amplitude dependent emittance growth in the horizontal but not in the vertical plane.
	\item[2017 experiment] Excitation amplitude dependent emittance growth in the vertical plane but not in the horizontal plane for an excitation in V and H+V. For an excitation only in H no emittance growth is observed.
\end{description}
Similar changes of the emittance as predicted in simulations were also experimentally observed as depicted in Fig.~\ref{fig:7thexp}, where emittance growth was observed only in the horizontal plane for pulsing in H in 2016, and only in the vertical plane for pulsing in V and H+V in 2017 with no emittance growth at all for pulsing in H. Just as for the $10^{\mathrm{th}}$~turn pulsing, the emittance growth can be divided also for the $7^{\mathrm{th}}$~turn pulsing in a phase of fast adjustment of the beam distribution and the reach of a new equilibrium state indicated in blue (fast adjustment) and black (equilibrium state) in Fig.~\ref{fig:7thexp}. Also in this case, the distribution changes are directly visible on the beam profiles taken with the BSRT (see Appendix~\ref{app:sec:7}, Fig.~\ref{fig:7thexpprof}).
\begin{figure*}[h]
		2016 experiment, $7^\mathrm{th}$ H	\\	\begin{minipage}[t]{0.32\linewidth}
		\centering
		\includegraphics[height=0.75\linewidth]{2016_bunch_intensity_h7th_no_damper_avg.png}
	\end{minipage}	
	\begin{minipage}[t]{0.32\linewidth}
		\centering
		\includegraphics[height=0.75\linewidth]{2016_emith_avg_rel_h7th_no_damper.png}
	\end{minipage}	
	\begin{minipage}[t]{0.32\linewidth}
		\centering
		\includegraphics[height=0.75\linewidth]{2016_emitv_avg_rel_h7th_no_damper.png}
	\end{minipage}	
		2017 experiment, $7^\mathrm{th}$ H+V\\
	\begin{minipage}[t]{0.32\linewidth}
		\centering
		\includegraphics[height=0.75\linewidth]{2017_bunch_intensity_hv7th_no_damper_avg.png}
	\end{minipage}	
	\begin{minipage}[t]{0.32\linewidth}
		\centering
		\includegraphics[height=0.75\linewidth]{2017_emith_avg_rel_hv7th_no_damper.png}
	\end{minipage}	
	\begin{minipage}[t]{0.32\linewidth}
		\centering
		\includegraphics[height=0.75\linewidth]{2017_emitv_avg_rel_hv7th_no_damper.png}
	\end{minipage}	
	\caption{\label{fig:7thexp} Relative horizontal (top) and vertical (bottom) emittance obtained during the 2016 experiments (left) and the 2017 experiment (right) averaged over the bunches experiencing the same excitation amplitude. For all bunches the damper was not active.}
\end{figure*}

@@

\subsection{random excitation\label{sec:simexran}}
The random excitation presents in general the pulsing pattern leading to the fastest halo depletion rates achievable with an HEL. On the other hand this pattern is also the most dangerous one for the proton beam core. The aim of this experiment was to test the effect on the core in comparison to the resonant excitation represented in the 2017 experiment by primarily the $7^{\mathrm{th}}$~turn pulsing as for $8^{\mathrm{th}}$~turn pulsing basically no effect was observed. The random excitation is in general so powerful as it excites all frequencies of the beam equally.
		no excitation
		\includegraphics[width=1.0\linewidth]{2017injnocolc15o+19_6noerru_dp0_amp.png}
		uniform random H+V
		\includegraphics[width=1.0\linewidth]{2017injnocolc15o+19_6noerruranadthv_1nrad_dp0_amp.png}
	\caption{\label{fig:ran2017fmaamp} FMA analysis in amplitude space without excitation (left) and for a uniform random excitation with 1~nrad amplitude (right) in H+V based on the 2017 injection optics with no machine errors and a tune of (62.27,60.295).}
This can be seen clearly on the example of the FMA analysis in amplitude space depicted in Fig.~\ref{fig:ran2017fmaamp}. In the case of a random excitation, the diffusion in tune is increased in comparison to the $k^{\mathrm{th}}$~turn pulsing rather equally for all particle amplitudes.
		\includegraphics[width=1.0\linewidth]{2017injerra2b2u_ranadt_3_5um_intensity.png}
		\includegraphics[width=1.0\linewidth]{2017injerra2b2u_ranadt_3_5um_sigm.png}
		\includegraphics[width=1.0\linewidth]{2017injerra2b2u_ranadt_3_5um_emit1.png}
		\includegraphics[width=1.0\linewidth]{2017injerra2b2u_ranadt_3_5um_emit2.png}
	\caption{\label{fig:ransim} Relative intensity (top left), bunch length (top right), and horizontal (bottom left) and vertical (bottom right) emittance obtained from distribution tracking based on the 2017 injection optics with standard errors and $(Q_x,Q_y)=(62.27,60.295)$. The solid line indicates the case with no excitation, and the dotted and dashed line the results for a uniform random excitation of 12~nrad and 24~nrad respectively.}
The distribution tracking shown in Fig.~\ref{fig:ransim} also reveals a quite symmetric evolution of the beam parameters for a random excitation:
\begin{itemize}
	\item Emittance growth is only caused by an excitation in the same plane.
	\item The emittance growth is quantitatively comparable for an excitation in only the corresponding plane. Only a small increase or decrease is observed if the excitation is applied in both planes. The difference could be due to coupling.
	\item The random excitation leads to an increase of the emittance growth rate. Compared to the $k^{\mathrm{th}}$~turn pulsing there is explicitly no initial adjustment phase of the distribution.
	\item Negligible losses are observed independent of the plane of excitation.
\end{itemize}
\begin{figure*}[b]
	random excitation V\\
	\begin{minipage}[t]{0.32\linewidth}
		\includegraphics[height=0.75\linewidth]{2017_bunch_intensity_vran_no_damper_avg.png}
	\begin{minipage}[t]{0.32\linewidth}
		\includegraphics[height=0.75\linewidth]{2017_emith_avg_rel_vran_no_damper.png}
	\begin{minipage}[t]{0.32\linewidth}
		\includegraphics[height=0.75\linewidth]{2017_emitv_avg_rel_vran_no_damper.png}
		random excitation H+V\\
	\begin{minipage}[t]{0.32\linewidth}
		\includegraphics[height=0.75\linewidth]{2017_bunch_intensity_hvran_no_damper_avg.png}
	\begin{minipage}[t]{0.32\linewidth}
		\includegraphics[height=0.75\linewidth]{2017_emith_avg_rel_hvran_no_damper.png}
	\begin{minipage}[t]{0.32\linewidth}
		\includegraphics[height=0.75\linewidth]{2017_emitv_avg_rel_hvran_no_damper.png}
	\caption{\label{fig:ranexp} Relative intensity (left) and relative horizontal (center) and vertical (right) emittance for a random excitation only in V (top) and in H+V (bottom). Emittance growth is only caused in the plane of excitation.}

Qualitatively, this behavior could be also confirmed experimentally during the 2017 experiments. Here excitation amplitude dependent emittance growth was only recorded in the plane of excitation with all excitation amplitude dependent emittance growth rates in a comparable range. As example the relative horizontal and vertical emittance for pulsing in V and in H+V are depicted in Fig.~\ref{fig:ranexp}. Note that as expected from simulations and despite the large excitation amplitude dependent emittance growth, the growth rates stays constant. There is explicitly no fast adjustment phase succeeded by a new equilibrium state as observed for the $7^{\mathrm{th}}$ and $10^{\mathrm{th}}$~turn pulsing.
\begin{figure}[h]
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		reference bunch,\\ no excitation
		\includegraphics[width=1.0\linewidth]{profile_v_ranv_slot_1698.png}
	\end{minipage}
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		excited bunch,\\ random excitation V
		\includegraphics[width=1.0\linewidth]{profile_v_ranv_slot_1532.png}
	\end{minipage}
	\caption{\label{fig:ranexpprof}Vertical beam profiles measured with the Beam Synchrotron Radiation Telescope (BSRT) during the 2017 experiments. The profiles are taken at the end of the random excitation in V. For both bunches the transverse damper is not active. The distribution changes in the reference bunch (left), here slot 1698, are much less than for the bunch experiencing the maximum excitation (right), here slot 2696.}
\end{figure}
The changes in emittance are also directly visible in the beam profiles measured with the BSRT. As an example Fig.~\ref{fig:ranexpprof} shows the profile for a random excitation only in V in comparison to a reference bunch. The distribution also stays Gaussian in case of a random excitation in contrast to the $10^{\mathrm{th}}$~turn pulsing in which case it clearly assumes a non-Gaussian shape (see Fig.~\ref{fig:10thexpprof}).

\begin{figure*}[h]
	\begin{minipage}[t]{0.32\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2017_scale_amp_ranh_lblshort.png}
	\end{minipage}	
	\begin{minipage}[t]{0.32\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2017_scale_amp_ranv_lblshort.png}
	\end{minipage}	
	\begin{minipage}[t]{0.32\linewidth}
		\centering
		\includegraphics[width=1.0\linewidth]{2017_scale_amp_ranhv_lblshort.png}
	\end{minipage}	
	\caption{\label{fig:ranexplossplane} Comparison of scaling of loss rates with excitation amplitude for a random excitation only in H (top), only in V (center) and in H+V (bottom) measured during the 2017 experiment. The excitation amplitude error is 15\% indicated by the horizontal error bars. The loss rate is considerably reduced for the bunches with transverse damper active and is approximately additive with the excitation plane, meaning that double the loss rate is observed for an excitation in H+V compared to only in H or only in V.}
\end{figure*}
The experimental results for a maximum excitation amplitude of $A_{\mathrm{max}}=1$~nrad and $A_{\mathrm{max}}=6$~nrad also confirm the in general small losses predicted in simulations. The losses then increase more dramatically around the time when the maximum excitation amplitude is increased to $A_{\mathrm{max}}=12$~nrad (see Fig.~\ref{fig:ranexpprof}). This is most likely the point when, due to the large emittance growth, the physical aperture starts cutting into the more central region of the distribution. Furthermore, the dependence of the loss rate on the excitation plane is additive, meaning that double the loss rate is observed for an excitation in H+V compared to only in H or only V. The loss rate is also considerably reduced for the bunches with transverse damper active independent of the plane the excitation is applied in (see Fig.~\ref{fig:ranexplossplane}).

\subsection{Effect of the transverse damper\label{sec:damp}}
\begin{figure}[b]
	\centering
	random excitation H+V\\
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		damper not active
		\includegraphics[width=1.0\linewidth]{2017_emitv_avg_rel_vran_no_damper.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		damper active
		\includegraphics[width=1.0\linewidth]{2017_emitv_avg_rel_vran_with_damper.png}
	\end{minipage}	
	\centering
	$7^{\mathrm{th}}$~turn H+V\\
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		damper not active
		\includegraphics[width=1.0\linewidth]{2017_emitv_avg_rel_v7th_no_damper_no_text.png}
	\end{minipage}	
	\begin{minipage}[t]{0.49\linewidth}
		\centering
		damper active
		\includegraphics[width=1.0\linewidth]{2017_emitv_avg_rel_v7th_with_damper_no_text.png}
	\end{minipage}	
	\caption{\label{fig:damp} Relative intensity for bunches with transverse damper not active (top) and transverse damper active (bottom) for $7^{\mathrm{th}}$~turn pulsing in V (left) and for a random excitation only in V (right).}
\end{figure}
For all $k^{\mathrm{th}}$~turn pulsing patterns tested experimentally, the transverse damper does in general not change any of the observables studied during the experiments, explicitly losses, emittance and beam distribution suggesting that indeed it is not capable of damping any resonant excitation. On the other hand, the transverse damper considerably decreases any changes of the above parameters in case of a random excitation. As an example, the vertical emittance growth for the $7^{\mathrm{th}}$~turn pulsing in V  is compared with a random excitation in V in Fig.~\ref{fig:damp} and the loss rates in Fig.~\ref{fig:ranexploss} illustrating impressively the aforesaid behavior. 
\textcolor{red}{Any idea why? Daniel, help!}

\section{Summary\label{sec:sum}}
In comparison to the $7^{\mathrm{th}}$~turn pulsing, the loss rate for a random excitation is smaller for a maximum excitation amplitude of $A_{\mathrm{max}}=6$~nrad, but larger for $A_{\mathrm{max}}=12$~nrad (see Fig.~\ref{fig:ranexploss}). This indicates an increase of the diffusion mostly in the beam tails for $7^{\mathrm{th}}$~turn pulsing, while for a random excitation mostly the core is affected.
\begin{figure}[b]
	\centering
	\includegraphics[width=0.8\linewidth]{2017_scale_amp_7th_ranhv_lblshort.png}
	\caption{\label{fig:ranexploss}  Comparison of scaling of loss rates with excitation amplitude for a random excitation and $7^{\mathrm{th}}$~turn pulsing in H+V measured during the 2017 experiment. The excitation amplitude error is 15\% indicated by the horizontal error bars.}
\end{figure}

This paper intends to give an overview of the effects of the HEL on the proton beam core and the experiments conducted in this context at the LHC in 2016~\cite{resexmd2016} and 2017~\cite{resexmd2017}. For a perfect HEL without any imperfections no effect on the core is to be expected as the field at the center vanishes by construction. Only in the case of imperfections in the guidance and profile of the electron beam, a residual kick on the proton beam core can occur. In this case and assuming HL-LHC parameters (Tables~\ref{tab:hllhc_param}--\ref{tab:hel_param}), the contribution from the electron lens bends with approximately $0.5$~nrad to first order is negligible with respect to the kick expected from the central region of the HEL of approximatley $20$~nrad to first order caused by profile imperfections in the electron beam.

@@

%\section{Additional plots for random excitation during 2017 experiments}
%\label{app:sec:ran}
